---
title: "Assignment 3"
author: "Thomas Soupionis"
date: "2025-03-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages
```{r}
library(tidyverse)
```

# Download Data
```{r} 
dow_data <- read_csv("../data/official_dowjones_data.csv")
spx_data <- read_csv("../data/official_spx_data.csv")
nasdaq_data <- read_csv("../data/official_nasdaq_data.csv")
```

# List all the columns/variables in the data
```{r}
names(spx_data)
names(dow_data)
names(nasdaq_data)
```

# Describing the Variables
## Dependent Variable
The dependent variable is the Return of the index. This variable measures the percentage change in the index value over a specified period. It represents the outcome of interest: whether an investment in the index fund yields a positive return over various time horizons.

## Independent Variable(s)
The key independent variable is the Time Horizon for the investment. This variable indicates the duration (e.g., 2 years, 5 years, 10 years) over which the return is calculated. It allows us to explore how the likelihood of a positive return changes with the length of the investment period. Additional independent variables might include market conditions, volatility measures, or economic indicators if available.

# Renaming Both Column for Each Dataset
```{r}
dow_data <- rename(dow_data,
                   Investment_Date = Date,
                   Closing_Price = Close)
nasdaq_data <- rename(nasdaq_data,
                   Investment_Date = Date,
                   Closing_Price = Close)
spx_data <- rename(spx_data,
                   Investment_Date = Date,
                   Closing_Price = Close)
```

# Verify Column Name Changes
```{r}
names(dow_data)
names(nasdaq_data)
names(spx_data)
```


