---
title: "Final Research Report"
author: "Thomas Soupionis"
date: "2025-05-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 0. Load R Packages
```{r}
# tidyverse for data wrangling and ggplot2 graphics, lubridate for date parsing
library(tidyverse)
library(lubridate)
library(ggplot2)
```
# 1. Research Question and Variable Definitions
**Research Question:** What are the distributions of 2-, 5-, and 10-year returns for the Dow Jones Industrial Average, S&P 500, and NASDAQ Composite, and how does holding period affect return consistency? Additionally, how do these outcomes differ when we limit the analysis to periods beginning at the S&P 500 inception (March 4, 1957)?

**Dependent Variable:**
- **Return_pct:** Numeric variable representing the percent return over a specified holding period (2, 5, or 10 years).

**Independent Variables:**
- **Horizon:** Categorical variable indicating holding period with three levels: “2 Years”, “5 Years”, and “10 Years”.
- **Index:** Categorical variable indicating the market index with three levels: “Dow Jones”, “S&P 500”, and “NASDAQ”.

# 2. Data Import and Preparation (Full Data)

**Data Source**: ^SPX, ^NDQ, and ^DJI on Stooq.com where all columns are deleted except Date and Close

```{r}
# Read daily closing data and add an "Index" identifier
# ymd() and mdy() convert character dates to Date objects

dow_data    <- read_csv("../data/official_dowjones_data.csv", col_types = cols(.default = col_guess(), Date = col_character())) %>%
  mutate(
    # parse both YYYY-MM-DD and M/D/YYYY date formats
    Date = parse_date_time(Date, orders = c("Y-m-d", "m/d/Y")),
    Index = "Dow Jones"
  )

spx_data    <- read_csv("../data/official_spx_data.csv") %>%
  mutate(
    Date = mdy(Date), 
    Index = "S&P 500")

nasdaq_data <- read_csv("../data/official_nasdaq_data.csv") %>%
  mutate(Date = mdy(Date), 
         Index = "NASDAQ")

# Function to compute percent return over 2-, 5-, and 10-year lags
# 504 trading days is about 2 years, 1260 is about 5 years, 2520 is about 10 years
calculate_returns <- function(df) {
  df %>%
    arrange(Date) %>%
    mutate(
      Return_2yr  = (Close / lag(Close, 504)  - 1) * 100,
      Return_5yr  = (Close / lag(Close, 1260) - 1) * 100,
      Return_10yr = (Close / lag(Close, 2520) - 1) * 100
    )
}

# Apply the function to each index series
dow_ret    <- calculate_returns(dow_data)
spx_ret    <- calculate_returns(spx_data)
nasdaq_ret <- calculate_returns(nasdaq_data)

# Combine all into one long data frame for plotting
all_returns <- bind_rows(dow_ret, spx_ret, nasdaq_ret) %>%
  select(Date, Index, starts_with("Return")) %>%
  pivot_longer(
    cols      = starts_with("Return"),
    names_to  = "Horizon",
    values_to = "Return_pct"
  ) %>%
  mutate(
    Horizon = recode(Horizon,
      Return_2yr  = "2 Years",
      Return_5yr  = "5 Years",
      Return_10yr = "10 Years"
    ),
    # set factor levels for correct order
    Horizon = factor(Horizon, levels = c("2 Years", "5 Years", "10 Years"))
  ) %>%
  filter(!is.na(Return_pct))

head(all_returns)
```

# 3. Data Import and Preparation (Post-SPX Inception)
Data Import and Preparation (Post-SPX Inception)
```{r}
# Restrict Dow to post S&P inception date
spx_start <- min(spx_data$Date)
dow_data_f <- dow_data %>% filter(Date >= spx_start)
# Recompute returns for filtered Dow
dow_ret_f <- calculate_returns(dow_data_f)

# Combine and reshape (filtered)
all_returns_filt <- bind_rows(dow_ret_f, spx_ret, nasdaq_ret) %>%
  select(Date, Index, starts_with("Return")) %>%
  pivot_longer(cols = starts_with("Return"),
               names_to = "Horizon", values_to = "Return_pct") %>%
  mutate(
    Horizon = recode(Horizon,
      Return_2yr  = "2 Years",
      Return_5yr  = "5 Years",
      Return_10yr = "10 Years"),
    Horizon = factor(Horizon, levels = c("2 Years", "5 Years", "10 Years"))
  ) %>%
  filter(!is.na(Return_pct))

head(all_returns_filt)
```

# 4. Summary Statistics on Full vs Filtered

```{r}
# Calculate mean, median, standard deviation, and positive/negative return percentages per horizon
get_summary <- function(data, label) {
  data %>% group_by(Horizon) %>%
    summarize(
      Scenario      = label,
      Mean_Return   = mean(Return_pct),
      Median_Return = median(Return_pct),
      SD_Return     = sd(Return_pct),
      Pct_Positive  = mean(Return_pct > 0) * 100,
      Pct_Negative  = mean(Return_pct < 0) * 100
    )
}

sum_full <- get_summary(all_returns, "Full Data")
sum_filt <- get_summary(all_returns_filt, "Post-1957")

# Combine and display
results <- bind_rows(sum_full, sum_filt)
knitr::kable(results, digits = 2,
             caption = "Summary Statistics for Full vs Post-SPX Inception Data")
```

## 4.1 Interpreting Summary Statistics
- **SD >= mean**: Returns are volatile relative to their average, indicating higher risk.
- **SD <= mean**: Returns cluster tightly around the mean, indicating lower risk.
- In both full and post-1957 data, the **2- and 5-year** horizons have SD > mean (higher short-term risk), while the **10-year** horizon has SD < mean (long-term investing reduces relative volatility).

# 5. Summary Statistics by Index
```{r}
stats_filt_idx <- all_returns_filt %>%
  group_by(Index, Horizon) %>%
  summarize(
    Mean_Return   = mean(Return_pct),
    Median_Return = median(Return_pct),
    SD_Return     = sd(Return_pct),
    Pct_Positive  = mean(Return_pct > 0) * 100,
    Pct_Negative  = mean(Return_pct < 0) * 100
  )
knitr::kable(stats_filt_idx, digits = 2,
             caption = "Summary Statistics by Index")
```
## 5.1 Interpreting Post-1957 Summary by Index
- **Dow Jones**: 2-year SD (22.2) > mean (16.2) indicates higher short-term volatility; at 5 years SD > mean (46.6 vs 45) shows moderate risk; at 10 years SD (94.7) < mean (110.1) reflects lower relative risk and more consistent long-term gains.  
- **NASDAQ**: 2-year SD (34.6) > mean (26.0) signals volatile 2-year periods; 5-year SD (74.3) < mean (80.0) suggests improved stability; 10-year SD (155.9) < mean (210) demonstrates strong, reliable long-term performance.  
- **S&P 500**: 2-year SD (22.8) > mean (17.7) shows short-term swings; 5-year SD (46.9) slightly < mean (48.7) indicates moderate volatility; 10-year SD (90) < mean (116.1) confirms reduced risk and dependable returns over a decade.  

# 6. Visualizing Distributions of Returns: Full vs Filtered
```{r fig.width=25, fig.height=15}
# Compute stats for full data
stats_full <- all_returns %>%
  group_by(Horizon) %>%
  summarize(
    Mean_Return   = mean(Return_pct),
    Median_Return = median(Return_pct),
    Pct_Positive  = mean(Return_pct > 0) * 100
  )

# Plot full data histograms
p1 <- ggplot(all_returns, aes(x = Return_pct)) +
  geom_histogram(bins = 50, fill = 'grey70', color = 'black') +
  geom_vline(data = stats_full, aes(xintercept = Mean_Return),
             color = 'red', linetype = 'dashed', linewidth = 0.7) +
  geom_text(data = stats_full,
            aes(x = Inf, y = Inf,
                label = sprintf("Mean=%.1f%%
Median=%.1f%%
%.0f%%+", 
                                Mean_Return, Median_Return, Pct_Positive)),
            hjust = 1.1, vjust = 1.1, size = 7) +
  facet_wrap(~ Horizon, scales = 'free_x') +
  labs(title = 'Full Data: Distribution of Returns') +
  theme_minimal()

# Compute stats for filtered data
stats_filt <- all_returns_filt %>%
  group_by(Horizon) %>%
  summarize(
    Mean_Return   = mean(Return_pct),
    Median_Return = median(Return_pct),
    Pct_Positive  = mean(Return_pct > 0) * 100
  )

# Plot filtered data histograms
p2 <- ggplot(all_returns_filt, aes(x = Return_pct)) +
  geom_histogram(bins = 50, fill = 'steelblue', color = 'black') +
  geom_vline(data = stats_filt, aes(xintercept = Mean_Return),
             color = 'red', linetype = 'dashed', linewidth = 0.7) +
  geom_text(data = stats_filt,
            aes(x = Inf, y = Inf,
                label = sprintf("Mean=%.1f%%
Median=%.1f%%
%.0f%%+", 
                                Mean_Return, Median_Return, Pct_Positive)),
            hjust = 1.1, vjust = 1.1, size = 7) +
  facet_wrap(~ Horizon, scales = 'free_x') +
  labs(title = 'Post-1957 Data: Distribution of Returns') +
  theme_minimal()

# Print side by side
library(gridExtra)
grid.arrange(p1, p2, ncol = 1)
```

## 6.1 Description of Distributions
### Full Data
- **2 Years:** Wide distribution with ~24% negative returns; mean ~18%, median ~17%.
- **5 Years:** Bulk shifts right with ~19% negative; mean ~51%, median ~45.5%.
- **10 Years:** Tight cluster with ~11% negative; mean ~119.5%, median ~99.5%.

### Post-1957 Data
- **2 Years:** Right-shifted, ~20% negative; mean ~19.5%, median ~18%.
- **5 Years:** ~17% negative; mean ~56%, median ~51%.
- **10 Years:** ~8% negative; mean ~140%, median ~127%.

# 7. Visualizing Relationship: Holding Period vs Return
```{r fig.width=12, fig.height=8}
# Boxplot:
#   • notch = approximate 95% CI around the median (±1.58·IQR/sqrt(n))
#   • red dot = mean
#   • y-axis limited to -50%–300% to focus on main distribution
all_compare <- bind_rows(
  all_returns %>% mutate(Scenario = 'Full'),
  all_returns_filt %>% mutate(Scenario = 'Post-1957')
)

ggplot(all_compare, aes(x = Horizon, y = Return_pct, fill = Scenario)) +
  geom_boxplot(notch = TRUE, outlier.shape = NA, position = position_dodge(width = 0.8)) +
  stat_summary(
    fun = mean, geom = "point",
    shape = 20, size = 3, color = "red",
    position = position_dodge(width = 0.8)
  ) +
  coord_cartesian(ylim = c(-50, 300)) +
  labs(
    title = 'Returns by Horizon: Full vs Post-1957',
    x = 'Holding Period', y = 'Return (%)'
  ) +
  theme_minimal() +
  theme(legend.position = 'none')
```

## 7.1 Explanation of Boxplots
- Notches show 95% confidence intervals around medians—non‑overlapping notches imply significantly different medians.
- Red dots mark mean returns.
- By capping the y-axis to -50%–300%, we focus on the core distribution and reduce the visual impact of extreme outliers.

Notice how the post‑1957 data consistently shows higher medians and means and tighter interquartile ranges, highlighting reduced relative volatility in the modern market.

# 8. Histograms by Index 
```{r fig.width=12, fig.height=12}
# Compute stats by index and horizon
stats_filt_idx <- all_returns_filt %>%
  group_by(Index, Horizon) %>%
  summarize(
    Mean_Return   = mean(Return_pct),
    Median_Return = median(Return_pct),
    Pct_Positive  = mean(Return_pct > 0) * 100
  )

# Post-1957 data by index with corner annotations
ggplot(all_returns_filt, aes(x = Return_pct)) +
  geom_histogram(bins = 50, fill = 'steelblue', color = 'black') +
  geom_vline(data = stats_filt_idx, aes(xintercept = Mean_Return),
             color = 'red', linetype = 'dashed', linewidth = 0.7) +
  geom_text(data = stats_filt_idx,
            aes(
              x = Inf, y = Inf,
              label = sprintf("Mean=%.1f%%
Median=%.1f%%
%.0f%%+", 
                              Mean_Return, Median_Return, Pct_Positive)
            ),
            hjust = 1.1, vjust = 1.1, size = 5) +
  facet_grid(Index ~ Horizon, scales = 'free_x') +
  labs(title = 'Post-1957 Data by Index: Distribution of Returns') +
  theme_minimal()
```

## 8.1. Interpreting Histograms by Index
- Across indices, the 2-year histograms show wider distributions and greater negative-return percentages, indicating shorter-term volatility.
- The 5-year panels shift right with fewer negatives, illustrating improved performance consistency mid-term.
- The 10-year panels cluster tightly with right-skewed tails, demonstrating strong long-term gains and minimal downside across all indices.
- NASDAQ exhibits the highest means and percentages positive in each horizon, reflecting its growth-oriented nature.

# 9. Boxplots by Index
```{r fig.width=15, fig.height=6}
ggplot(all_returns_filt, aes(x = Horizon, y = Return_pct, fill = Horizon)) +
  geom_boxplot(notch = TRUE, outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point",
               shape = 20, size = 3, color = "red") +
  facet_wrap(~ Index, nrow = 1) +
  coord_cartesian(ylim = c(-50, 300)) +
  labs(
    title = 'Post-1957: Returns by Horizon for Each Index',
    x = 'Holding Period',
    y = 'Return (%)'
  ) +
  theme_minimal() +
  theme(legend.position = 'none')
```

## 9.1. Interpreting Boxplots by Index
- The notched boxplots confirm median differences across horizons within each index: longer horizons have higher medians.
- Red mean points lie above medians for longer horizons, especially pronounced in NASDAQ’s 10-year box, indicating skew.
- Interquartile ranges shrink with holding period, showing reduced volatility over longer terms across all indices.
- NASDAQ’s boxes consistently sit higher than Dow and S&P, reaffirming its superior return profile across horizons.

# 10. Statistical Tests for Horizon and Index
```{r}
# ANOVA: difference in mean returns across holding periods (Horizon)
anova_horizon_full <- aov(Return_pct ~ Horizon, data = all_returns)
anova_horizon_filt <- aov(Return_pct ~ Horizon, data = all_returns_filt)

# ANOVA: difference in mean returns across indices (Index)
anova_index_full <- aov(Return_pct ~ Index, data = all_returns)
anova_index_filt <- aov(Return_pct ~ Index, data = all_returns_filt)

# Summaries of tests
summary(anova_horizon_full)
summary(anova_horizon_filt)
summary(anova_index_full)
summary(anova_index_filt)
```

*These ANOVA tests evaluate whether mean returns differ significantly by holding period (Horizon) and by market index (Index) for both the full and post-SPX inception datasets.*

## 10.1. Explanation of Statistical Tests
The ANOVA results show highly significant differences in mean returns:

- **Horizon (Full Dataset):** F(2, 179871) = 28435, p < 2e-16 indicates return means differ across 2-, 5-, and 10-year holding periods for the full data.
- **Horizon (Post-SPX Dataset):** F(2, 131214) = 27282, p < 2e-16 likewise confirms significant mean differences post-1957.
- **Index (Full Dataset):** F(2, 179871) = 5401, p < 2e-16 shows significant differences in return means among Dow, S&P 500, and NASDAQ for the full data.
- **Index (Post-SPX Dataset):** F(2, 131214) = 3114, p < 2e-16 confirms the same for the post-1957 subset.

All p-values are below conventional significance levels (***), indicating strong evidence that both holding period and market index affect average returns.

# 11. Annualized Return Analysis 

To compare these multi‐year returns to typical bank or CD rates, we convert each holding‐period return into an **annualized geometric return**:

```{r echo=FALSE}
all_ann <- all_returns_filt %>%
  mutate(
    years = case_when(
      Horizon == "2 Years"  ~ 2,
      Horizon == "5 Years"  ~ 5,
      Horizon == "10 Years" ~ 10
    ),
    Annualized = (1 + Return_pct/100)^(1/years) - 1
  )
```

# 12. Annualized Summary Statistics by Horizon
```{r}
ann_horizons <- all_ann %>%
  group_by(Horizon) %>%
  summarize(
    Mean_Ann      = mean(Annualized)*100,
    Median_Ann    = median(Annualized)*100,
    SD_Ann        = sd(Annualized)*100,
    Pct_Positive  = mean(Annualized > 0) * 100,
    Pct_Negative  = mean(Annualized < 0) * 100
  )

knitr::kable(
  ann_horizons,
  digits = 2,
  caption = "Table 3: Annualized Return Statistics by Horizon (Post-1957)"
)
```

## 12.1: Interpreting Annualized Returns by Horizon
- **2 Years:** Mean = 8.63%, Median = 8.72%, SD = 12.47%; 80.23% of periods are positive, 19.75% negative, indicating shorter horizons retain some downside risk.
- **5 Years:** Mean = 8.23%, Median = 8.57%, SD = 7.68%; 83.29% positive, 16.70% negative, showing improved consistency relative to 2-year returns.
- **10 Years:** Mean = 7.97%, Median = 8.54%, SD = 5.37%; 91.79% positive, 8.21% negative, demonstrating highest reliability and lowest volatility over a decade.

# 13. Annualized Summary Statistics by Index
```{r}
ann_index <- all_ann %>%
  group_by(Index, Horizon) %>%
  summarize(
    Mean_Ann      = mean(Annualized)*100,
    Median_Ann    = median(Annualized)*100,
    SD_Ann        = sd(Annualized)*100,
    Pct_Positive  = mean(Annualized > 0) * 100,
    Pct_Negative  = mean(Annualized < 0) * 100
  )

knitr::kable(
  ann_index,
  digits = 2,
  caption = "Table 4: Annualized Return Statistics by Index (Post-1957)"
)
```

## 13.1. Interpreting Annualized Returns by Index
- **Dow Jones:** 2-year (Mean=7.33%, SD=10.30%, 78.12% positive), 5-year (Mean=6.88%, SD=6.61%, 81.82% positive), 10-year (Mean=6.69%, SD=4.97%, 88.93% positive) – consistent improvement and reduced volatility with longer horizons.
- **NASDAQ:** 2-year (Mean=11.14%, SD=16.08%, 80.50% positive), 5-year (Mean=10.99%, SD=9.26%, 86.80% positive), 10-year (Mean=10.76%, SD=5.63%, 95.06% positive) – highest returns and reliability over all horizons.
- **S&P 500:** 2-year (Mean=7.96%, SD=10.77%, 82.15% positive), 5-year (Mean=7.42%, SD=6.71%, 82.03% positive), 10-year (Mean=7.11%, SD=4.74%, 92.15% positive) – moderate returns with strong long-term consistency.

# 14. Takeaway
The analysis demonstrates that **holding period** and **market index** both play significant roles in U.S. equity returns:

- **Short-term horizons (2–5 years)** carry elevated volatility and a meaningful chance of negative outcomes (15–24%); median returns hover near the average, but risk remains high relative to mean performance.  
- **Long-term horizons (10 years)** deliver substantially higher cumulative returns (110–210% raw; 7.97–10.76% annualized) with far lower relative volatility (SD < mean) and negative-return probabilities under 12% (around 8% annualized), affirming the benefits of patience.  
- **Post-1957 era** consistently boosts mean and median returns across all horizons and indices, reflecting a more stable, growth-oriented market after S&P 500 inception.  
- **Index comparisons** reveal NASDAQ outperforms in both absolute and annualized returns, especially over longer periods, while Dow and S&P 500 show similar patterns of risk reduction but at lower average returns.

**Context vs. Traditional Savings Vehicles:** Compared to typical cash and fixed-income options, these equity returns are substantial:
- **Certificates of Deposit (CDs):** Often yield under 1%–2% annualized.  
- **High-Yield Savings Accounts:** Currently offer around 3%–4% APY.  
- **Money Market Funds:** Typically yield around 4% APY.  

By contrast, even the lowest annualized equity returns (approximately 7.97% over 10 years post-1957) significantly outpace these alternatives, demonstrating the historical advantage of a long-term equity strategy for young investors seeking growth beyond traditional safe-haven rates.

**Implications for Younger Investors:** For those balancing risk and return—perhaps new savers choosing between savings accounts, CDs, or equities—the data suggest that a diversified, long-horizon equity allocation can offer materially higher returns over time, while still mitigating downside through extended holding periods. 
